A Spectral Analysis of the Domain Decomposed Monte Carlo Method for Linear Systems

### 2015-02-05 doc/ned2015/slattery_wilson_evans.tex
formatting figures in ned paper

An alternative approach to approximate matrix inversion is to employ Monte
Carlo methods that sample a distribution with an expectation value equivalent
to that of the inverted operator. Such methods have been in existence for
decades with the earliest reference noted here an enjoyable manuscript
published in 1950 by Forsythe and Leibler \cite{forsythe_matrix_1950}. In
their outline, Forsythe and Leibler in fact credit the creation of this
technique to J. Von Neumann and S.M. Ulam some years earlier than its
publication. In 1952 Wasow provided a more formal explanation of Von Neumann
and Ulam's method \cite{wasow_note_1952} and Hammersley and Handscomb's 1964
monograph \cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
\cite{spanier_monte_1969} present additional detail on a method adjoint to
that described by Neumann and Ulam. To improve their convergence, Halton
leveraged the Monte Carlo algorithm in an iterative refinement scheme where a
residual Monte Carlo solution serves as a correction to a fixed point
iteration \cite{halton_sequential_1962}. In recent years, the adjoint
Neumann-Ulam method has been leveraged in Monte Carlo Synthetic Acceleration
(MCSA) methods that are competitive with conventional methods including Krylov
methods and are applicable to a wide variety of linear systems
\cite{evans_monte_2009,evans_monte_2014}. To advance the applicability of
these new MCSA methods, we therefore aim to apply the underlying Monte Carlo
methods to large-scale problems in parallel computing environments.

@@

To test the relationships developed by the spectral analysis, we form
two simple numerical experiments using the diffusion model problem:
one to measure the length of the random walks as a function of the
iteration matrix eigenvalues, and one to measure the domain leakage
fraction as a function of the iteration matrix eigenvalues and the
discretization properties. Before doing this, we verify our
computation of the spectral radius of the iteration matrix by
numerically computing the largest eigenvalue of the diffusion operator
using an iterative eigenvalue solver. For this verification, a $100
\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
absorption cross varied from 0 to 100 while the scattering cross
section was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
the measured spectral radius of the iteration matrix and the computed
spectral radius for the preconditioned diffusion operator using
Eq~(\ref{eq:iteration_radius}) as function of the absorption to
scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
observed between the analytic and numerical results with all data
points computed within the tolerance of the iterative eigenvalue
solver.
\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=4in,clip]{measured_spec_rad.png}
  \end{center}
  \caption{Measured and analytic preconditioned diffusion operator
    spectral radius as a function of the absorption cross section to
    scattering cross section ratio. Values of $h=0.01$, $h=0.1$, and
    $h=1.0$ were used. The red data was computed numerically by an
    eigensolver while the black dashed data was generated by
    Eq~(\ref{eq:iteration_radius}).}
  \label{fig:measured_spec_rad}
\end{figure}

@@

With the spectral data in hand, we can go about setting up an
experiment to measure the length of the random walks generated by the
adjoint Neumann-Ulam solver. To do this, we again use a $100 \times
100$ square grid with $h=0.1$ and the absorption cross varied from 0
to 100 while the scattering cross section was fixed at unity. Three
weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were
used with 10,000 histories generated by a point source of strength 1
in the center of the domain. For each of the histories, the number of
transitions made was tallied to provide an effective value of $k$ for
each history. This value was then averaged over all histories to get a
measured value of $k$ for the particular operator. On the left,
Figure~\ref{fig:measured_length} presents these measurements as well
as the analytic result computed by Eq~(\ref{eq:analytic_k}) as a
function of the iteration matrix spectral radius, $\rho(\ve{H})$. On
the right, Figure~\ref{fig:measured_length} gives the relative error
between the predicted and observed results. We note good qualitative
agreement between the measured and analytic results. However, we
observe a larger relative error for both long and short random walks.
\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=6.5in,clip]{measured_length.png}
  \end{center}
  \caption{Measured and analytic random walk length as a function of
    the iteration matrix spectral radius. The weight cutoff was
    varied with \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8}. In the left
    plot, the red data was computed numerically by an adjoint
    Neumann-Ulam implementation while the black dashed data was
    generated by Eq~(\ref{eq:analytic_k}). In the right plot, the
    relative error between the predicted and measured results is
    presented for each weight cutoff.}
  \label{fig:measured_length}
\end{figure}

@@

To compute the leakage fraction numerically, \sn{3}{5} histories were
sampled from a uniform source of strength unity over the global
domain. At the start of a stage of histories, the number of histories
starting in the center domain was computed and as the stage
progressed, the number of histories that exited that domain was
tallied with the ratio of the two numbers providing a numerical
measure for the leakage fraction. Figure~\ref{fig:measured_leakage}
gives the domain leakage measurements for the domain in the center of
the global grid as well as the analytic result computed by
Eqs~(\ref{eq:wigner_domain_leakage}) and
(\ref{eq:mean_chord_domain_leakage}) as a function of the iteration
matrix spectral radius.
\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=4.5in,clip]{leakage_variation.png}
  \end{center}
  \caption{Measured and analytic domain leakage as a function of the
    iteration matrix spectral radius. To test the behavior with
    respect to domain size, $n_i=50$, $n_i=100$,and $n_i=200$ were
    used. The red data was computed numerically by a
    domain-decomposed adjoint Neumann-Ulam implementation, the black
    dashed data was generated by
    Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
    approximation, and the dashed-dotted black data was generated by
    Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner rational
    approximation.}
  \label{fig:measured_leakage}
\end{figure}
Again, we note good qualitative agreement between the measured and
analytic quantities but we begin to see the limits of the leakage
approximations. To compare the quality of the two approximations, the
absolute error between the computed leakage fraction and that
generated by the Wigner rational and mean chord approximations is
plotted in Figure~\ref{fig:leakage_error} for all domain sizes
tested. 
\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=4.5in,clip]{leakage_error.png}
  \end{center}
  \caption{Measured and analytic domain leakage absolute error as a
    function of the iteration matrix spectral radius.  To test the
    behavior with respect to domain size, $n_i=50$ (green),
    $n_i=100$ (blue), and $n_i=200$ (red) were used. The dashed
    lines represent the error using the Wigner rational
    approximation while the solid lines represent the error using
    the mean-chord approximation.}
  \label{fig:leakage_error}
\end{figure}
From these error results, the mean chord approximation is shown to
have a lower error for ill-conditioned systems as compared to the
Wigner approximation while the Wigner approximation produces less
error for more well-conditioned systems. We also note that for the
optically thick domains, the error is likely corresponded to that
observed in Figure~\ref{fig:measured_length} for the $k$ parameter
while the large relative error in $k$ for optically thin domains does
not affect the approximation significantly. In general, the mean chord
approximation is a better choice to estimate the leakage fraction in a
domain from the adjoint Neumann-Ulam method and except for a single
data point with $n_i=50$, the mean chord approximation yielded leakage
fractions within 0.05 of the measured results. As the domain becomes
more optically thick (with both increasing $n_i$ and decreasing
$\rho(\ve{H})$), the approximations are more accurate.

>>>>

An alternative approach to approximate matrix inversion is to employ Monte
Carlo methods that sample a distribution with an expectation value equivalent
to that of the inverted operator. Such methods have been in existence for
decades with the earliest reference noted here an enjoyable manuscript
published in 1950 by Forsythe and Leibler \cite{forsythe_matrix_1950}. In
their outline, Forsythe and Leibler in fact credit the creation of this
technique to J. Von Neumann and S.M. Ulam some years earlier than its
publication. In 1952 Wasow provided a more formal explanation of Von Neumann
and Ulam's method \cite{wasow_note_1952} and Hammersley and Handscomb's 1964
monograph \cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
\cite{spanier_monte_1969} present additional detail on a method adjoint to
that described by Neumann and Ulam. To improve their convergence, Halton
leveraged the Monte Carlo algorithm in an iterative refinement scheme where a
residual Monte Carlo solution serves as a correction to a fixed point
iteration \cite{halton_sequential_1962}. In recent years, the adjoint
Neumann-Ulam method has been leveraged in Monte Carlo Synthetic Acceleration
(MCSA) methods that, for certain classes of problems, are competitive with
conventional methods including Krylov methods and are applicable to a wide
variety of linear systems \cite{evans_monte_2009,evans_monte_2014}. To advance
the applicability of these new MCSA methods, we therefore aim to apply the
underlying Monte Carlo methods to large-scale problems in parallel computing
environments.

@@

To test the relationships developed by the spectral analysis, we form
two simple numerical experiments using the diffusion model problem:
one to measure the length of the random walks as a function of the
iteration matrix eigenvalues, and one to measure the domain leakage
fraction as a function of the iteration matrix eigenvalues and the
discretization properties. Before doing this, we verify our
computation of the spectral radius of the iteration matrix by
numerically computing the largest eigenvalue of the diffusion operator
using an iterative eigenvalue solver. For this verification, a $100
\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
absorption cross varied from 0 to 100 while the scattering cross
section was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
the measured spectral radius of the iteration matrix and the computed
spectral radius for the preconditioned diffusion operator using
Eq~(\ref{eq:iteration_radius}) as function of the absorption to
scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
observed between the analytic and numerical results with all data
points computed within the tolerance of the iterative eigenvalue
solver.
\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=\textwidth]{measured_spec_rad.png}
  \end{center}
  \caption{Measured and analytic preconditioned diffusion operator
    spectral radius as a function of the absorption cross section to
    scattering cross section ratio. Values of $h=0.01$, $h=0.1$, and
    $h=1.0$ were used. The red data was computed numerically by an
    eigensolver while the black dashed data was generated by
    Eq~(\ref{eq:iteration_radius}).}
  \label{fig:measured_spec_rad}
\end{figure}

@@

With the spectral data in hand, we can go about setting up an
experiment to measure the length of the random walks generated by the
adjoint Neumann-Ulam solver. To do this, we again use a $100 \times
100$ square grid with $h=0.1$ and the absorption cross varied from 0
to 100 while the scattering cross section was fixed at unity. Three
weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were
used with 10,000 histories generated by a point source of strength 1
in the center of the domain. For each of the histories, the number of
transitions made was tallied to provide an effective value of $k$ for
each history. This value was then averaged over all histories to get a
measured value of $k$ for the particular operator. On the left,
Figure~\ref{fig:measured_length} presents these measurements as well
as the analytic result computed by Eq~(\ref{eq:analytic_k}) as a
function of the iteration matrix spectral radius, $\rho(\ve{H})$. On
the right, Figure~\ref{fig:measured_length} gives the relative error
between the predicted and observed results. We note good qualitative
agreement between the measured and analytic results. However, we
observe a larger relative error for both long and short random walks.
\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=\textwidth]{measured_length.png}
  \end{center}
  \caption{Measured and analytic random walk length as a function of
    the iteration matrix spectral radius. The weight cutoff was
    varied with \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8}. In the left
    plot, the red data was computed numerically by an adjoint
    Neumann-Ulam implementation while the black dashed data was
    generated by Eq~(\ref{eq:analytic_k}). In the right plot, the
    relative error between the predicted and measured results is
    presented for each weight cutoff.}
  \label{fig:measured_length}
\end{figure}

@@

To compute the leakage fraction numerically, \sn{3}{5} histories were
sampled from a uniform source of strength unity over the global
domain. At the start of a stage of histories, the number of histories
starting in the center domain was computed and as the stage
progressed, the number of histories that exited that domain was
tallied with the ratio of the two numbers providing a numerical
measure for the leakage fraction. Figure~\ref{fig:measured_leakage}
gives the domain leakage measurements for the domain in the center of
the global grid as well as the analytic result computed by
Eqs~(\ref{eq:wigner_domain_leakage}) and
(\ref{eq:mean_chord_domain_leakage}) as a function of the iteration
matrix spectral radius.
\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=\textwidth]{leakage_variation.png}
  \end{center}
  \caption{Measured and analytic domain leakage as a function of the
    iteration matrix spectral radius. To test the behavior with
    respect to domain size, $n_i=50$, $n_i=100$,and $n_i=200$ were
    used. The red data was computed numerically by a
    domain-decomposed adjoint Neumann-Ulam implementation, the black
    dashed data was generated by
    Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
    approximation, and the dashed-dotted black data was generated by
    Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner rational
    approximation.}
  \label{fig:measured_leakage}
\end{figure}
Again, we note good qualitative agreement between the measured and
analytic quantities but we begin to see the limits of the leakage
approximations. To compare the quality of the two approximations, the
absolute error between the computed leakage fraction and that
generated by the Wigner rational and mean chord approximations is
plotted in Figure~\ref{fig:leakage_error} for all domain sizes
tested. 
\begin{figure}[ht!]
  \begin{center}
    \includegraphics[width=\textwidth]{leakage_error.png}
  \end{center}
  \caption{Measured and analytic domain leakage absolute error as a
    function of the iteration matrix spectral radius.  To test the
    behavior with respect to domain size, $n_i=50$ (green),
    $n_i=100$ (blue), and $n_i=200$ (red) were used. The dashed
    lines represent the error using the Wigner rational
    approximation while the solid lines represent the error using
    the mean-chord approximation.}
  \label{fig:leakage_error}
\end{figure}
From these error results, the mean chord approximation is shown to
have a lower error for ill-conditioned systems as compared to the
Wigner approximation while the Wigner approximation produces less
error for more well-conditioned systems. We also note that for the
optically thick domains, the error is likely corresponded to that
observed in Figure~\ref{fig:measured_length} for the $k$ parameter
while the large relative error in $k$ for optically thin domains does
not affect the approximation significantly. In general, the mean chord
approximation is a better choice to estimate the leakage fraction in a
domain from the adjoint Neumann-Ulam method and except for a single
data point with $n_i=50$, the mean chord approximation yielded leakage
fractions within 0.05 of the measured results. As the domain becomes
more optically thick (with both increasing $n_i$ and decreasing
$\rho(\ve{H})$), the approximations are more accurate.
